Analysis Approach
========================================================
```{r libraries, echo = FALSE}
library(MASS)
library(ggplot2)
library(gvlma)
library(car)
library(lattice)
```

```{r setoptions, echo = FALSE}
opts_chunk$set(echo = TRUE, warning = FALSE)
#opts_chunk$set(dev = 'pdf')
```

```{r}
# Downloading the data
data(mtcars)

#Analyzing data set
str(mtcars)
summary(mtcars)

#Searching for missing values
table(is.na(mtcars))

#creating tidy data set
tidy <- mtcars

#CREATE a matrix scatter plot to show the relationships and indentify which are factor variables
scatterplotMatrix(tidy, spread=FALSE, lty.smooth=2,
                                     main="Scatter Plot Matrix")
```

From the above analysis, factor variables are:
1. $cyl
2. $vs
3. $am
4. $gear
5. $carb


```{r cleaningDataSet}
# creating factor variables
tidy$am <- as.factor(mtcars$am)
tidy$cyl <- as.factor(mtcars$cyl)
tidy$gear <- as.factor(mtcars$gear)
tidy$carb <- as.factor(mtcars$carb)
tidy$vs <- as.factor(mtcars$vs)

tidy$am <- factor(mtcars$am, levels=c(0,1),
labels=c("Automatic", "Manual"))
```



```{r}
# Analyzing if automatic vs manual is better for MPG
boxplot(mpg~am, data=tidy, xlab="Transmission Type", ylab="Miles per Gallon (MPG)",
main="MGP by Transmission Type")
```

```{r}
meanTransmission <- with(tidy, tapply(mpg,am, FUN=mean))
print(meanTransmission)
```

We see that miles per gallon is better as estimate = `r meanTransmission[2]`.


```{r, fig.align='center', fig.height = 11, fig.width=11}
fit1 <- stepAIC(lm(data=tidy,mpg~.,direction="both",test="F"))
summary(fit1)
fit2 <- lm(formula = mpg ~ cyl + hp + am + wt, data = tidy)
summary(fit2)
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(fit2)
```

Hence, let's add interactions terms     


```{r}
fit3 <- lm(formula = mpg ~ cyl + hp + am + wt + wt*am, data = tidy)
summary(fit3)
anova(fit2,fit3)
```



```{r}
# Removing hp
fit4_1 <- lm(formula = mpg ~ cyl + wt + am, data = tidy)
summary(fit4_1)

# Removing hp and cyl
fit4_2 <- lm(formula = mpg ~ wt + am, data = tidy)
summary(fit4_2)

fit4_3 <- lm(formula = mpg ~ cyl + wt + am + wt*am, data = tidy)
summary(fit4_3)


anova( fit2,fit3, fit4_1, fit4_2, fit4_3)
```

Hence its model 4_3 that we choose!

```{r, echo=FALSE}

g <-ggplot(tidy, aes(x=wt, y=mpg, shape=am, color = cyl )) + geom_point()
print(g)

````
## Let's analyze outliers

```{r}
cutoff <- 4/(nrow(tidy)-length(fit4_2$coefficients))
plot(fit4_3, which=4, cook.levels=cutoff)
abline(h=cutoff, col="red")

```

```{r}
influencePlot(fit4_3, id.method="noteworthy", main="Influence Plot",
sub="Circle size is proportional to Cook's distance")
```


```{r}
removingVal <- rownames(tidy) %in% c("Toyota Corolla", "Fiat 128", "Chrysler Imperial", "Maserati Bora")
tidy2<- tidy
tidy2$Cars <- rownames(tidy2)
tidy2 <- tidy[!removingVal,]
#print(tidy2)
fit5 <- lm(formula = mpg ~ cyl + wt + am + wt*am , data = tidy2)
summary(fit5)
plot(fit5)

#anova(fit4_3, fit5) 
```

```{r}
#Doing further residual analysis - Normality
qqPlot(fit5, labels=row.names(tidy2), id.method="identify",
simulate=TRUE, main="Q-Q Plot")
# Look at verbage around Q-Q plot when writing this up p.193
```
Decided not to remove Lincoln as it's not that big of a circle

```{r}
# residplot <- function(fit, nbreaks=10) {
# z <- rstudent(fit)
# hist(z, breaks=nbreaks, freq=FALSE,
# xlab="Studentized Residual",
# main="Distribution of Errors")
# rug(jitter(z), col="brown")
# curve(dnorm(x, mean=mean(z), sd=sd(z)),
# add=TRUE, col="blue", lwd=2)
# lines(density(z)$x, density(z)$y,
# col="red", lwd=2, lty=2)
# legend("topright",
# legend = c( "Normal Curve", "Kernel Density Curve"),
# lty=1:2, col=c("blue","red"), cex=.7)
# }
# residplot(fit5)
```

```{r}
# INDEPENDENCE OF ERRORS
durbinWatsonTest(fit5)
```
The nonsignificant p-value suggests a lack of autocorrelation, and conversely an independence of errors.


```{r}
# HOMOSCEDASTICITY
ncvTest(fit5)
spreadLevelPlot(fit5)
```

```{r}
# Global Validation of the Model
gvmodel<- gvlma(fit5)
summary(gvmodel)
```

```{r}
sqrt(vif(fit5))
```

There could be a VIF problem for am